Radiocarbon measurement of CO~2~ from carbonate minerals using a gas ion source with an open split interface

### 2022-01-11 doc/AMS-15/AMS-15_paper.Rmd
Add figures, more text, some things from poster.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

The NOSAMS USAMS is an AMS system based on a 3MV Tandetron accelerator. A NEC 134-sample gas-capable MCSNICS source. Analysis of samples as CO2 simplifies sample preparation, making analysis less expensive and faster.

@@

Most interfaces to gas-accepting sputter ion sources are a closed inlet design. A mixture of sample and helium is introduced to the source through a capillary, and flow is controlled by the pressure in the sample volume and the length and diameter of the capillary. Flow may be adjusted by changing the pressure by varying the sample volume. We designed a simpler open inlet where a capillary samples gas in a flowing stream that exhausts to the atmosphere. This defines the delivery pressure as 1 ATM and therefore flow is controlled by the capillary and is not adjustable. The flow of CO2 to the source may be varied by controlling the ratio of CO2 to helium in the sample stream. An open inlet has the advantage of simplicity; flow of CO2 to the source is preset and there are no separate sample transfer and dilution steps. A disadvantage of the open split is that the flow of CO2 to the source varies with sample size (CO2:He) and changes over the course of a measurement due to dillution of the sample by displacement helium. Salazar and Szidat, Roberts gasbench, other open split methods.

@@

Figure 1. photo and drawing of inlet.

@@

@@

Carbonate samples are weighed into 7 ml septa-sealed Vacutainer vials (Labco # 768W). Sealed vials are evacuated, then 1 ml 85% H3PO4 is added. The samples are allowed to react at 60C for 4-6 h. Sample pressure was adjusted to 1 ATM by adding helium or removing sample using a syringe.

@@

@@

A new target was used for each sample. Targets were allowed to "burn in" for 5-10 minutes with pure helium delivered to the source. Initial currents of > 10 uA dropped rapidly, eventually settling to 100-400 nA. After a stable base current was reached, the sample vial was installed on the double needle, starting sample flow to the source. Sample current rises to a stable level within 30 s, at which time, data acquisition is started. 10 blocks of 90 s for a total of 15 m.

@@

Setting flow to the source to 10uL/min provides excess CO2 to source until the helium dilution ratio increases to > 1:1. Another approach may be to backfill vials with helium and run at a higher dilution ratio and flow to the source. This should produce a smaller change in He:CO2 over the course of a run. MICADAS runs at 95% He:CO2, which would mean producing < 0.5mL CO2 in a 7mL vial and backfilling with helium before displacing to the source. Some intermediate dilution may be appropriate. Alternatively, vials could be flushed with helium prior to acidification, rather than evacuating, leading to some concentration of CO2 in He at slightly > 1 ATM.

@@

First tests with carbonate in vials were hit-or-miss. Problems with capillary clogging and low currents led to poor results for several samples. A pair of NOSAMS2 samples produced good currents, results agreed well, but were about 3% higher than expected Fm. A second set of carbonates produced better currents (~5uA) and better results. Agreement of replicates varied from 3 to 16 permil. Near-modern NOSAMS2 was very close to the consensus value, while C-2 and C-1 were elevated by 2 and 3.6% respectively, indicating that a blank correction is needed.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Abstract

@@

Radiocarbon Accelerator Mass Spectrometry (AMS) is a sensitive method for determining the age, provenance, and composition of carbon-containing materials. A primary disadvantage of the method is its cost, which often limits the number of samples analyzed. This forces researchers to carefully choose samples to be run and limits repeat measurements for error determination. One driver of the cost of AMS analysis is the time and effort required for sample preparation, including reduction of the sample to graphite. Analysis of analyte CO2 directly via a gas-accepting AMS ion source eliminates the cost and effort of graphitization and returns results more rapidly. The NOSAMS USAMS is an AMS system based on a 3MV Tandetron accelerator. A NEC 134-sample gas-capable MCSNICS source. Analysis of samples as CO2 simplifies sample preparation, making analysis less expensive and faster.

@@

Most interfaces to gas-accepting sputter ion sources are a closed inlet design. A mixture of sample and helium is introduced to the source through a capillary, and flow is controlled by the pressure in the sample volume and the length and diameter of the capillary. Flow may be adjusted by changing the pressure by varying the sample volume. We designed a simpler open inlet where a capillary samples gas in a flowing stream that exhausts to the atmosphere. This defines the delivery pressure as 1 ATM and therefore flow is controlled by the capillary and is not adjustable. The flow of CO2 to the source may be varied by controlling the ratio of CO2 to helium in the sample stream. An open inlet has the advantage of simplicity; flow of CO2 to the source is preset and there are no separate sample transfer and dilution steps. A disadvantage of the open split is that the flow of CO2 to the source varies with sample size (CO2:He) and changes over the course of a measurement due to dilution of the sample by displacement helium. Salazar and Szidat, Roberts gasbench, other open split methods.

@@

@@

![Experimental setup. Helium and standard gas regulators at left, sample vial and open split at top right.](images/gas_control.jpg)

Our method uses an open split and glass capillary to control gas flow to the ion source. Compared to closed inlet systems which regulate introduction of a CO2/Helium mixture to the ion source by regulating sample pressure, our method is simpler and faster, but does not maintain a constant flow of CO2 to the source. We compensate for this by limiting CO2 flow to a range that produces relatively constant C- ion current and stable ratios.

![Inlet layout. Helium displacement gas is delivered to vial through a double needle, mixed sample and helium flows to an open split, and a glass capillary delivers gas to the ion source.](images/HGIS_open_split_drawing.svg)

@@

Carbonate samples are weighed into 7 ml septa-sealed Vacutainer vials (Labco \# 768W). Sealed vials are evacuated, then 1 ml 85% H3PO4 is added. The samples are allowed to react at 60C for 4-6 h. Sample pressure was adjusted to 1 ATM by adding helium or removing sample using a syringe.

@@

1.  Weigh 10-30mg Carbonate in septa-sealed 7mL vials produces 2-6mL CO<sub>2</sub> \@ \~1ATM
2.  Evacuate and Acidify with 1 mL H3PO4 using automated Gilson prep bench
3.  After reaction is complete, add helium to bring vials to ambient pressure.

![A modified liquid handling system is used to evacuate and add acid to sample vials.](images/gilson.png)

### Measurement

1.  Displace gas to open split with 100 uL/min He.
2.  Introduce mixture to source \@ 30 uL/min
3.  Acquire data: 90s x 10 measurements.

-   Normalizing standards: live tank gas in vials
-   Blanks: C-1 or TIRI-F in vials

### Data reduction

Normalization to standard, blank correction, and estimation of measurement error are done with the following steps:

1.  Correct for fractionation using online AMS <sup>13</sup>C measurement.
2.  Determine per-sample mean and error (repeatabilty and counting error).
3.  Normalize to mean of Gas standard.
4.  Blank correct: Blank subtraction using mean of carbonate blanks

Errors from normalization and blank correction are propagated at each step. The per-sample error agrees well with variability of replicates, and measurement of known-value standards shows that the method has no significant bias.

Data reduction is done using an R package maintained on github (<https://github.com/blongworth/hgis>).

@@

A new target was used for each sample. Targets were allowed to "burn in" for 5-10 minutes with pure helium delivered to the source. Initial currents of \> 10 uA dropped rapidly, eventually settling to 100-400 nA. After a stable base current was reached, the sample vial was installed on the double needle, starting sample flow to the source. Sample current rises to a stable level within 30 s, at which time, data acquisition is started. 10 blocks of 90 s for a total of 15 m.

@@

Setting flow to the source to 10uL/min provides excess CO2 to source until the helium dilution ratio increases to \> 1:1. Another approach may be to backfill vials with helium and run at a higher dilution ratio and flow to the source. This should produce a smaller change in He:CO2 over the course of a run. MICADAS runs at 95% He:CO2, which would mean producing \< 0.5mL CO2 in a 7mL vial and backfilling with helium before displacing to the source. Some intermediate dilution may be appropriate. Alternatively, vials could be flushed with helium prior to acidification, rather than evacuating, leading to some concentration of CO2 in He at slightly \> 1 ATM.

@@

First tests with carbonate in vials were hit-or-miss. Problems with capillary clogging and low currents led to poor results for several samples. A pair of NOSAMS2 samples produced good currents, results agreed well, but were about 3% higher than expected Fm. A second set of carbonates produced better currents (\~5uA) and better results. Agreement of replicates varied from 3 to 16 permil. Near-modern NOSAMS2 was very close to the consensus value, while C-2 and C-1 were elevated by 2 and 3.6% respectively, indicating that a blank correction is needed.
